### 2022-03-30 ehoh2/ehoh2.tex
more

%
The $\lambda$-superposition calculi were
previously implemented in Zipperposition, and
extensive experiments with various heuristic choices have been performed
(Chapter \ref{ch:ho-techniques}). In \ehohii{}'s implementation, we used
these experiences to choose a set of effective rules that could easily be
retrofitted into an originally first-order prover. Another principle that guided 
the design of \ehohii{} was \emph{gracefulness}: we made sure that our changes
do not impact the strong first-order performance of E and $\lambda$-free higher-order performance of Ehoh.

@@

We \NumberOK{evaluated} \ehohii{} on a selection of proof assistants benchmarks
as well as all higher-order theorems in the TPTP library \cite{gs-17-tptp}
(Sect.~\ref{sec:ehoh2:eval}). We found
that \ehohii{} clearly outperforms Ehoh on all benchmarks. It outperformed all other higher-order provers on the
proof assistant benchmarks; on the TPTP benchmarks it ended up second only 
to the cooperative version of Zipperposition, which employs Ehoh as a
backend. An arguably fairer comparison without the backend puts \ehohii{} in
first place for both benchmark suites.
We also compared the performance of \ehohii{} with E on first-order
%%% @PETAR: I made this slightly sentence shorter and more forceful. Please
%%% check if you agree. --JB
problems and found that little overhead has been introduced by the
extension to higher-order logic.

@@

Ehoh represents $\lambda$-free higher-order terms using flattened, spine notation
(Sect.~\ref{sec:ehoh:types-and-terms}).
Thus, the terms $\cst{f}$, $\cst{f} \, \cst{a}$, and $\cst{f} \, \cst{a} \,
\cst{b}$ are represented by the cells $\cst{f}$,
$\cst{f}(\cst{a})$, and $\cst{f}(\cst{a}, \cst{b})$, respectively.
To ensure free variables are
perfectly shared, Ehoh treats applied free variables differently: arguments are
not applied directly to a free variable, but using an internal symbol
\internalat{} of variable arity. For example, the term $X \, \cst{a} \, \cst{b}$ is
represented by the cell $\internalat(X, \cst{a}, \cst{b})$. This ensures that
two different occurrences of the free variable $X$ correspond to the same object,
which makes substitutions more efficient.
%%% @PETAR(read): This tries to explain an invisible problem. Either we need to explain
%%% it better, or we just omit it (my favorite option). It's old Ehoh stuff anyway. --JB
%Yet, to
%efficiently access variable arguments we do not use binary application symbol,
%but flattened representation.

@@

%%% @PETAR(read): Sounds odd---we wouldn't be extending the term representation of an
%%% already HO prover. --JB
%Unlike other higher-order provers, most of which are written in functional
%programming languages,
Extending the term representation of Ehoh with a new term
kind involves intricate manipulation of the cell data structure. De Bruijn
indices must be represented as other cells with either negative or positive
\texttt{f\_code}. However, this has to be done in such a way that a De Bruijn
index can never be mistaken for a function symbol or a variable.

@@

To avoid this, we use Ehoh's perfectly shared
terms and their \texttt{properties} field.
%
We introduce the property \texttt{HasBetaReducibleSubterm}, which is set if
%%% @PETAR(is is good): "is" or "may be" below? (Overapproximation?) --JB
a cell is $\beta$-reducible.
Whenever a new cell that contains a
$\beta$-reducible term as a direct subterm is shared, the property is set.
Setting of the property is inductively continued when further superterms are
shared. For example, in the term $t = \cst{f} \, \cst{a} \, (\cst{g} ((\lambda x.\,
x)\,\cst{a}))$, the cells for $(\lambda x.\, x)\,\cst{a}$,
$\cst{g}\,((\lambda x.\, x)\,\cst{a})$ and $t$ itself have the property
\texttt{HasBetaReducibleSubterm} set.
%
When it needs to find $\beta$-reducible subterms, \ehohii{} will visit only the
cells with this property set. This further means that on first-order
subterms, a single bit masking operation is enough to determine that no subterm
should be visited.

@@

        \If{$\vn{n\_branch\_iter} \not= \fc{BindEnd}$ }
          \State push pair $(\vn{lhs},\vn{rhs})$ back to \vn{iter.constraints}
          \State push quadruple $(\vn{iter.constraints}, \vn{n\_branch\_iter}, $
          \State                  $\quad \vn{iter.steps}, \vn{iter.subst})$ onto $\vn{iter.bt\_state}$
          \State extend $\vn{iter.subst}$ with $\vn{n\_binding}$
          \State $\vn{iter.steps} \gets \vn{n\_steps}$
          \State $\vn{iter.branch\_iter} \gets \cn{BindBegin}$
        \ElsIf{\vn{lhs.head = rhs.head}}
          \State create constraint pairs of arguments of $\vn{lhs}$ and $\vn{rhs}$
          \State $\quad$ and push them to \vn{iter.constrants}
          \State $\vn{iter.branch\_iter} \gets \cn{BindBegin}$
        \EndIf
      \EndIf
    \ElsIf{$\vn{lhs.head} = \vn{rhs.head}$}
      \State create constraint pairs of arguments of $\vn{lhs}$ and $\vn{rhs}$
      \State $\quad$ and push them to \vn{iter.constrants}
    \Else{} \State $\vn{forward} \gets \fc{BacktrackIter(\vn{iter})}$
    \EndIf
  \EndWhile
  \State \Return $\vn{forward}$
  \EndFunction

@@

\looseness=-1
\fc{ForwardIter} begins by backtracking if the previous attempt was successful (i.e., 
all constraints were solved). If it finds a state from which it can continue,
it takes term pairs from $\vn{constraints}$ until
there are no more constraints or it is determined that no unifier exists. The terms
are normalized by instantiating the head variable with its binding and then
reducing the potential top-level $\beta$-redex that appears. This instantiation
and reduction process is repeated
until there are no more top-level $\beta$-redexes and the head is
not a variable bound to some term. Then the term with shorter
$\lambda$ prefix is expanded (only on the top-level) so that $\lambda$ prefixes have the
same length. Finally, the $\lambda$ prefix is ignored, and we
focus only on the body. In this way, we avoid fully substituting
and normalizing terms and perform just enough operations
to determine the next step of the procedure.

@@

We have modified the $\lambda$-free higher-order matching algorithm to treat
$\lambda$ prefixes as above in the unification procedure---by bringing the prefixes
to the same length and ignoring them afterwards. This ensures that the
algorithm will never try to match a free variable to a $\lambda$-abstraction,
making sure that $\beta$-redexes never appear (as in original $\lambda$-free
higher-order matching). We also modified the algorithm to ensure that free variables
are never bound to terms that have loose bound variables. This algorithm cannot
find many complex matching substitutions (matchers), but it can
efficiently determine whether two terms are variable renamings of each other
or whether a simple matcher can be used, as in the case of $(X \, (\lambda x. \,x) \,
\cst{b}, \cst{f} \, (\lambda x. \,x) \, \cst{b})$, where ${X \mapsto \cst{f}}$ is
usually the desired matcher. If this algorithm does not find a matcher and both
terms are patterns, pattern matching is tried.

@@

  \ourpara{Extensionality} \ehohii{} takes a pragmatic approach to reason about
  functional and Boolean extensionality: It uses \emph{abstracting} rules (\infname{Abs} rules of Sect.~\ref{sect:bool:native}) 
  which simulate basic superposition calculus rules,
  but do not require unifiability of the partner terms in the inference. More
  precisely, assume a core inference needs to be performed between two
  $\beta$-reduced terms $u$ and $v$, such that they can be represented as
  $u=C[s_1, \ldots, s_n]$ and $v=C[t_1, \ldots, t_n]$, where $C$ is the most general
  common context (more precisely, green common context that does not go into the structure of $\lambda$-abstraction or applied variables \cite{bbtvw-21-sup-lam}) of $u$ and $v$, not all of
  $s_i$ and $t_j$ are free variables, and for at least one $i$, $s_i \not= t_i$,
  $s_i$ and $t_i$ are not (applied) free variables, and they are of Boolean or
  function type. Then, the conclusion is formed by taking the conclusion $D$ of
  the core inference rule (which would be created if $s$ and $t$ are unifiable)
  and adding literals $s_1 \noteq t_1 \llor \cdots \llor s_n \noteq t_n$.

@@

  Counterintuitively, the higher-order versions of both provers outperform
  the first-order counterparts. However, the differerence is so small that it can
  be attributed to the changes to memory layout that affect the order in which
  clauses are chosen. Similar effects are seen when comparing the first-order versions.
  We would expected the less heavily modified version of E, Ehoh, to perform better,
  but due to subtle effects \ehohii{} wins.

>>>>

%
The $\lambda$-superposition calculi were
previously implemented in Zipperposition, and
extensive experiments with various heuristic choices have been performed
(Chapter \ref{ch:ho-techniques}). In \ehohii{}'s implementation, we used
these experiences to choose a set of effective rules that could easily be
retrofitted into an originally first-order prover. Another principle that guided 
the design of \ehohii{} was \emph{gracefulness}: We made sure that our changes
do not impact the strong first-order performance of E and $\lambda$-free higher-order performance of Ehoh.

@@

We \NumberOK{evaluated} \ehohii{} on a selection of proof assistants benchmarks
as well as all higher-order theorems in the TPTP library \cite{gs-17-tptp}
(Sect.~\ref{sec:ehoh2:eval}). We found
that \ehohii{} clearly outperformed Ehoh on all benchmarks. It outperformed all other higher-order provers on the
proof assistant benchmarks; on the TPTP benchmarks it ended up second only 
to the cooperative version of Zipperposition, which employs Ehoh as a
backend. An arguably fairer comparison without the backend puts \ehohii{} in
first place for both benchmark suites.
We also compared the performance of \ehohii{} with E on first-order
%%% @PETAR: I made this slightly sentence shorter and more forceful. Please
%%% check if you agree. --JB
problems and found that little overhead has been introduced by the
extension to higher-order logic.

@@

Ehoh represents $\lambda$-free higher-order terms using flattened, spine notation
(Sect.~\ref{sec:ehoh:types-and-terms}).
Thus, the terms $\cst{f}$, $\cst{f} \, \cst{a}$, and $\cst{f} \, \cst{a} \,
\cst{b}$ are represented by the cells $\cst{f}$,
$\cst{f}(\cst{a})$, and $\cst{f}(\cst{a}, \cst{b})$, respectively.
To ensure free variables are
perfectly shared, Ehoh treats applied free variables differently: Arguments are
not applied directly to a free variable, but using an internal symbol
\internalat{} of variable arity. For example, the term $X \, \cst{a} \, \cst{b}$ is
represented by the cell $\internalat(X, \cst{a}, \cst{b})$. This ensures that
two different occurrences of the free variable $X$ correspond to the same object,
which makes substitutions more efficient.
%%% @PETAR(read): This tries to explain an invisible problem. Either we need to explain
%%% it better, or we just omit it (my favorite option). It's old Ehoh stuff anyway. --JB
%Yet, to
%efficiently access variable arguments we do not use binary application symbol,
%but flattened representation.

@@

%%% @PETAR(read): Sounds odd---we wouldn't be extending the term representation of an
%%% already HO prover. --JB
%Unlike other higher-order provers, most of which are written in functional
%programming languages,
Extending the term representation of Ehoh with a new term
kind involves intricate manipulation of the cell data structure. De Bruijn
indices must be represented as other cells with either a negative or a positive
\texttt{f\_code}. However, this has to be done in such a way that a De Bruijn
index can never be mistaken for a function symbol or a variable.

@@

To avoid this, we use Ehoh's perfectly shared
terms and their \texttt{properties} field.
%
We introduce the property \texttt{HasBetaReducibleSubterm}, which is set if
%%% @PETAR(is is good): "is" or "may be" below? (Overapproximation?) --JB
a cell is $\beta$-reducible.
Whenever a new cell that contains a
$\beta$-reducible term as a direct subterm is shared, the property is set.
Setting of the property is inductively continued when further superterms are
shared. For example, in the term $t = \cst{f} \, \cst{a} \, (\cst{g} ((\lambda x.\,
x)\,\cst{a}))$, the cells for $(\lambda x.\, x)\,\cst{a}$,
$\cst{g}\,((\lambda x.\, x)\,\cst{a})$, and $t$ itself have the property
\texttt{HasBetaReducibleSubterm} set.
%
When it needs to find $\beta$-reducible subterms, \ehohii{} will visit only the
cells with this property set. This further means that on first-order
subterms, a single bit masking operation is enough to determine that no subterm
should be visited.

@@

        \If{$\vn{n\_branch\_iter} \not= \fc{BindEnd}$ }
          \State push pair $(\vn{lhs},\vn{rhs})$ back to \vn{iter.constraints}
          \State push quadruple $(\vn{iter.constraints}, \vn{n\_branch\_iter}, $
          \State                  $\quad \vn{iter.steps}, \vn{iter.subst})$ onto $\vn{iter.bt\_state}$
          \State extend $\vn{iter.subst}$ with $\vn{n\_binding}$
          \State $\vn{iter.steps} \gets \vn{n\_steps}$
          \State $\vn{iter.branch\_iter} \gets \cn{BindBegin}$
        \ElsIf{\vn{lhs.head = rhs.head}}
          \State push the constraint pairs of arguments of $\vn{lhs}$ and $\vn{rhs}$
                 to \vn{iter.constraints}
          \State $\vn{iter.branch\_iter} \gets \cn{BindBegin}$
        \EndIf
      \EndIf
    \ElsIf{$\vn{lhs.head} = \vn{rhs.head}$}
    \State push the constraint pairs of arguments of $\vn{lhs}$ and $\vn{rhs}$
           to \vn{iter.constraints}
    \Else{} \State $\vn{forward} \gets \fc{BacktrackIter(\vn{iter})}$
    \EndIf
  \EndWhile
  \State \Return $\vn{forward}$
  \EndFunction
  \vspace{2\jot}
  \pagebreak[2]

@@

\looseness=-1
\fc{ForwardIter} begins by backtracking if the previous attempt was successful (i.e., 
all constraints were solved). If it finds a state from which it can continue,
it takes term pairs from $\vn{constraints}$ until
there are no more constraints or it is determined that no unifier exists. The terms
are normalized by instantiating the head variable with its binding and then
reducing the potential top-level $\beta$-redex that appears. This instantiation
and reduction process is repeated
until there are no more top-level $\beta$-redexes and the head is
not a variable bound to some term. Then the term with shorter
$\lambda$ prefix is expanded (only on the top level) so that $\lambda$ prefixes have the
same length. Finally, the $\lambda$ prefix is ignored, and we
focus only on the body. In this way, we avoid fully substituting
and normalizing terms and perform just enough operations
to determine the next step of the procedure.

@@

We have modified the $\lambda$-free higher-order matching algorithm to treat
$\lambda$ prefixes as above in the unification procedure---by bringing the prefixes
to the same length and ignoring them afterwards. This ensures that the
algorithm will never try to match a free variable with a $\lambda$-abstraction,
making sure that $\beta$-redexes never appear (as in original $\lambda$-free
higher-order matching). We also modified the algorithm to ensure that free variables
are never bound to terms that have loose bound variables. This algorithm cannot
find many complex matching substitutions (matchers), but it can
efficiently determine whether two terms are variable renamings of each other
or whether a simple matcher can be used, as in the case of $(X \, (\lambda x. \,x) \,
\cst{b}, \cst{f} \, (\lambda x. \,x) \, \cst{b})$, where ${X \mapsto \cst{f}}$ is
usually the desired matcher. If this algorithm does not find a matcher and both
terms are patterns, pattern matching is tried.

@@

  \begin{sloppypar}
    \ourpara{Extensionality} \ehohii{} takes a pragmatic approach to reasoning about
    functional and Boolean extensionality: It uses \emph{abstracting} rules (\infname{Abs} rules of Sect.~\ref{sect:bool:native}) 
    which simulate basic superposition calculus rules,
    but do not require unifiability of the partner terms in the inference. More
    precisely, assume a core inference needs to be performed between two
    $\beta$-reduced terms $u$ and $v$, such that they can be represented as
    $u=C[s_1, \ldots, s_n]$ and $v=C[t_1, \ldots, t_n]$, where $C$ is the most general
    common context (more precisely, green common context that does not go into the structure of $\lambda$-abstraction or applied variables \cite{bbtvw-21-sup-lam}) of $u$ and $v$, not all of
    $s_i$ and $t_j$ are free variables, and for at least one $i$, $s_i \not= t_i$,
    $s_i$ and $t_i$ are not (possibly applied) free variables, and they are of Boolean or
    function type. Then, the conclusion is formed by taking the conclusion $D$ of
    the core inference rule (which would be created if $s$ and $t$ are unifiable)
    and adding literals $s_1 \noteq t_1 \llor \cdots \llor s_n \noteq t_n$.

  \end{sloppypar}

@@

  Counterintuitively, the higher-order versions of both provers outperform
  the first-order counterparts. However, the difference is so small that it can
  be attributed to the changes to memory layout that affect the order in which
  clauses are chosen. Similar effects are seen when comparing the first-order versions.
  We would expected the less heavily modified version of E, Ehoh, to perform better,
  but due to subtle effects \ehohii{} wins.

### 2022-03-30 satfol/satfol.tex
unif
biblio
more

\begin{defi}
   \label{def:definition}
   \looseness=-1
   Let $G$ be a clause set, $\cst{p}$ be a predicate symbol, and
   $\tupleempty{x}$ be distinct variables.
%
   The set $G$ is a \emph{definition set} for $\cst{p}$ if
   \begin{enumerate}
      \item $\cst{p}$ is singular for $G$
      \item $G$ consists of clauses of the form $\arbpredlit{\cst{p}(\tupleempty{x})} \cor C'$ (up to variable renaming)
      \item the variables in $C'$ are all among $\tupleempty{x}$
      \item all clauses in $\withpredpos{G}{\cst{p}}
      \flatres_{\!\cst{p}} \withpredneg{G}{\cst{p}}$ are tautologies
      \item $E(\tupleempty{\cst{c}})$ is unsatisfiable, where
      the \emph{environment} $E(\tupleempty{x})$ consists of all subclauses $C'$ of any
      $\arbpredlit{\cst{p}(\tupleempty{x})} \mathbin{\cor} C' \in G$ and
      $\tupleempty{\cst{c}}$ is a tuple of distinct fresh constants substituted
      in for $\tupleempty{x}$
   \end{enumerate}
\end{defi}

@@

Hidden-literal-based elimination, predicate elimination, \begin{qle}quasipure
elimination, \end{qle}and blocked clause
elimination all admit efficient implementations in a superposition
prover. In this section, we describe how to implement the first
\begin{noqle}two\end{noqle}\begin{qle}three\end{qle} sets of
techniques. For BCE, we refer to Kiesl et al.\ \cite{ksstb-2017-blockedfol}.
All techniques have been implemented in the Zipperposition prover
\cite{sc-supind-17}.

@@

\newcommand{\vn}[1]{\ensuremath{\mathit{#1}}} %variable name
\algrenewcommand\alglinenumber[1]{{\color{gray} \footnotesize #1}}
\algrenewcommand\algorithmicindent{0.75em}

>>>>

\begin{defi}
   \label{def:definition}
   \looseness=-1
   Let $G$ be a clause set and $\cst{p}$ be a predicate symbol.
%
   The set $G$ is a \emph{definition set} for $\cst{p}$ if
   \begin{enumerate}
      \item $\cst{p}$ is singular for $G$
      \item $G$ consists of clauses of the form
      $\arbpredlit{\cst{p}(\tupleempty{x})} \cor C'$ (up to variable renaming),
      where the variables $\tupleempty{x}$ are distinct
      \item the
      variables in $C'$ are all among $\cst{p}$'s arguments $\tupleempty{x}$
      \item all clauses in $\withpredpos{G}{\cst{p}}
      \flatres_{\!\cst{p}} \withpredneg{G}{\cst{p}}$ are tautologies
      \item $E(\tupleempty{\cst{c}})$ is unsatisfiable, where
      the \emph{environment} $E(\tupleempty{x})$ consists of all subclauses $C'$ of any
      $\arbpredlit{\cst{p}(\tupleempty{x})} \mathbin{\cor} C' \in G$ and
      $\tupleempty{\cst{c}}$ is a tuple of distinct fresh constants substituted
      in for $\tupleempty{x}$
   \end{enumerate}
\end{defi}

@@

Hidden-literal-based elimination, predicate elimination, \begin{qle}quasipure
elimination, \end{qle}and blocked clause
elimination all admit efficient implementations in a superposition
prover. In this section, we describe how to implement the first
\begin{noqle}two\end{noqle}\begin{qle}three\end{qle} sets of
techniques. For BCE, we refer to Kiesl et al.\ \cite{ksstb-2017-blockedfol}.
All techniques have been implemented in the Zipperposition prover.

@@

\algrenewcommand\alglinenumber[1]{{\color{gray} \footnotesize #1}}
\algrenewcommand\algorithmicindent{0.75em}
